### 2023-08-01 Manuscript/Draft2.tex
incorporating liz and tim's edits

The continental shelf off the coast of the Northeast United States spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species, making commercial fisheries an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which supported over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.). Maintaining a healthy ecosystem is therefore vital to sustained ecosystem health and economic prosperity of the region.

The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf. The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each strata in approximately the same 3-4 week time period in each season.

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process. Common model-based approaches include General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM).

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. The changes have resulted in a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area is an additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate change on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends. We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.

@@

We construct spatial population models for Yellowtail Flounder (\emph{Limanda ferruginea}), Atlantic Cod (\emph{Gadus morhua}), and Haddock (\emph{Melanogrammus aeglefinus}) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock. Movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. To address our research question we formulate the following inputs for the MFS package.

@@

We obtained a shapefile for the 15 strata that comprise Georges Bank, where strata are partitioned based primarily on depth and secondarily by latitude (Politis et al. 2014). The region was discritized into a raster with 88 rows and 144 columns to use as our modeling environment. Each cell in our simulation domain represents approximately 8.7 km\(^2\). A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters. Each of the species being modeled has multiple biologically distinct stocks along the Atlantic coast resulting from local environmental conditions. As a result, each stock inhabits a different number of strata on Georges Bank. Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata. Figure \ref{fig:strata-plot} shows the strata used in our model for each species.

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Individual terms in the formulation account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics. A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.

@@

Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022). The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure seasonal influences to allow the niche model to instead infer static habitat preferences independent of temperature. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we rescaled the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.

@@

\caption{Visual representation of niche model for Atlantic Cod.}\label{fig:hab-plot1}
\end{figure}

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. We consider increasing, decreasing, and constant population trends for Yellowtail for completeness. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

\caption{True population trends used in indexing analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
\end{figure}

\emph{Simulating Bottom Trawl Survey and Population Indexing}

After completing all simulations, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited strata twice each year. We sample each strata in the same weeks in which the spring and fall surveys take place (weeks 13 \& 14 in the spring and 37 \& 38 in the fall). The number of the samples taken reflect true target values for each strata and sampling cells were randomly selected.

An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extend of the population. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were removed from certain stratified random surveys for each species. We then use the biomass collected from our samples in contemporary abundance indexing methods to estimate population trends. Knowing the true population values in our simulations allows us to compute and compare the absolute error calculated from each estimation method.

The design-based method we used is the stratified mean, which divides the inhabited domain into \(N\) disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude. The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is relative to the given area. The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated by through the weighted average

\[
  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},$}
\]

@@

We compare stratified mean estimates to those derived from the model-based VAST approach. VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model that can be represented by

\[
  \makebox[\linewidth]{$p_*(i)= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
\]

for \(*=1,2\), where more specific functional forms of each component are further described in (Thorson 2019). VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved. As a result, VAST models take on the order of hours to complete.

@@

Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b). After running models for each of our scenarios using these settings, we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed our VAST model results and suggested we also consider the options shown in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). According to Jim, the Settings A values are representative of what stock assessment scientists typically use in their analyses. As a result we will focus our analysis on VAST results obtained using Settings A, but present Settings B results to demonstrate how the settings used in model-based approaches impact model performance.

Each indexing scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our abundance estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall errors and standard deviations. Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39. These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower errors than when covariates are included). We analyze these results further in the sections that follow.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes. A representative example of this trend for each species is shown on the log scale in Figure \ref{fig:RatioTrend}.

Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed similar patterns.

When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in turn decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

>>>>

The continental shelf off the coast of the Northeast United States spans from the Outer Banks of North Carolina to the Gulf of Maine. The region covers over 250,000 km\(^2\) of ocean, extending over 200 km from shore in the largest areas in New England to just 30 km off shore in the southern regions. This ecologically diverse region contains approximately 18,000 vertebrate marine species, making commercial fisheries an important part of local economies for centuries. In 2019, New England fisheries produced \$22 billion in sales, which supported over 200,000 jobs ({``{Fisheries Economics of the United States: Data and Visualizations}''} n.d.). Maintaining a healthy ecosystem is therefore vital to the economic prosperity of the region.

The Northeast Fisheries Science Center (NEFSC) has conducted a bottom trawl survey since 1963 to support assessment and management of the fish and invertebrate populations in the region (Azarovitz 1981; Politis et al. 2014). The survey uses a stratified random design where bottom trawl sampling takes place in predefined strata along the eastern continental shelf, where depth and latitude define the strata. The survey has created a rich time series data set with many uses including species-specific habitat identification, age and life history information, analysis of how environmental conditions influence species abundance, and estimating yearly species abundance trends to help inform stock assessments, and ultimately fishing quota limits. The survey takes place twice each year- once in the spring and again in the fall. Most spatial analyses and projections of future distributions typically assume a constant survey catchability and availability over time. For this reason, NOAA's survey design includes sampling each stratum in approximately the same 6 week time period in each season.

One component of a stock assessment is an estimation of yearly abundance that is often derived from the bottom trawl survey, when applicable. Stock assessment scientists choose from a number of approaches to obtain abundance estimates ranging from traditional design-based estimates to model-based estimates that vary in complexity. Design-based estimators rely on the design of the sampling scheme with the underlying assumption that the data being collected is representative of the population of interest. These methods do not account for spatial variation in samples and are not able to relate environmental influences with survey values. Model-based abundance estimates use statistical models to measure the relationship between response variables (such as presence or abundance) and predictor variables (such as environmental factors). Model-based estimators help account for complex relationships between variables and can help overcome problems related to the sampling process. Common model-based approaches include Generalized Linear Models (GLM), Generalized Additive Model (GAM), and General Linear Mixed Models (GLMM).

Due to a combination of climate change and shifts in circulation, the Northeast United States continental shelf has experienced rapid warming in recent decades. At the same time a number of recent studies have found a shift in spatial distributions of many species (Nye et al. 2009; Henderson et al. 2017; Kleisner et al. 2017). Since stock assessment models rely on accurate descriptions of population dynamics and contemporary patterns of spatial abundance, there is concern that rapid undocumented changes in spatial distributions of species will bias future stock assessments. More specifically, as fish populations shift their distributions over time, catchability and/or availability in the survey could change, altering the relationship between the index and the true population (Arreguı́n-Sánchez 1996; Langan et al. 2021). A species shifting its range beyond the survey area or outside of the assumed stock boundary are additional compounding factor to consider. Existing research has focused on temperature as the driver of such changes (Klein et al. 2017) and evidence suggests that failing to account for the impact of climate-induced change can lead to management challenges (Kerr et al. 2022). In these scenarios, management strategy evaluations have shown that misconceptions of stock status can lead to unintended overfishing, which can ultimately have detrimental ecologic and economic impacts (Mazur et al. 2023). We are therefore interested in analyzing the impact of climate driven movement and distribution changes on the accuracy of abundance estimates derived from NOAA's ongoing bottom-trawl survey along the east coast.

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depends on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends. We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the stratum specific mean catch per tow, which are then weighted by stratum area to obtain an overall mean. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). VAST also allows users to include covariate data to better inform the model. Covariates can be static (e.g.~habitat preferences) or dynamics (e.g.~temperature). We explore whether including environmental predictors leads to improved abundance estimates, which is particularly relevant as climate change progresses.

@@

We construct spatial population models for Yellowtail Flounder (\emph{Limanda ferruginea}), Atlantic Cod (\emph{Gadus morhua}), and Haddock (\emph{Melanogrammus aeglefinus}) on Georges Bank, hereafter referred to as Yellowtail, Cod, and Haddock. Movement of each species combines static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated spatial data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation data to mimic the bottom trawl survey and use the samples to compare the ability of contemporary index standardization methods to track population trends.

@@

We use the R package \emph{MixFishSim} (MFS) to model our populations (Dolder et al. 2020). MFS is a discrete spatiotemporal simulation tool that allows users to model multiple species under varying environmental conditions. The package uses a delay-difference population model with discrete processes for growth, death, and recruitment of the population. To address our research question we specify the following inputs for the MFS package.

@@

A shapefile for the 15 strata that comprise Georges Bank was discretized into a raster with 88 rows and 144 columns to use as our modeling environment. Each cell in our simulation domain represents approximately 8.7 km\(^2\). A fish stock is considered to be a subpopulation of a species that has similar intrinsic parameters. Each of the species being modeled has multiple biologically distinct stocks along the Atlantic coast resulting from local environmental conditions. As a result, each stock inhabits a different number of strata on Georges Bank. Haddock inhabit all 15 strata in the domain, Cod populate 13 strata, and Yellowtail can be found in 9 strata (Figure \ref{fig:strata-plot}).

@@

The time step for our models is one week. MFS uses a modified two-stage Deriso-Schnute delay difference equation that models the biomass in each cell in our study area (Dolder et al. 2020). Parameters in the population model account for growth of mature adults, mortality (natural and fishing), and the addition of new recruits. Recruitment is a function of the adult biomass that existed in the previous year and is added to the population incrementally throughout each species' predefined spawning period. Parameter inputs were either obtained from the literature or chosen to produce desired model dynamics, and therefore, these results cannot be compared directly with stock assessment results. A full list of parameters used in our model can be seen in Tables 1 and \ref{tab:paramsSCENARIOS}.

@@

The habitat input for each species in MFS should represent spatial habitat preferences for each species that are time-invariant and are instead related to static geographic features. Species-specific habitat preferences were derived from niche models for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022). The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and by representing a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure seasonal influences to allow the niche model to instead infer static habitat preferences independent of temperature. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we rescaled the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.

@@

\caption{Visual representation of niche model for Atlantic Cod. In the left most plot, tows with positive Cod catch are shown in red while tows without Cod are in blue. Positive tows are weighted by the biomass of the given catch. The middle two plots show our covariate predictors used in the *lrren* niche model- depth and median sediment size. The plot on the right shows the final niche model that has been rescaled between 0 and 1, where blue values are closer to 0 and red values are closer to 1.}\label{fig:hab-plot1}
\end{figure}

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Cod has seen significant decline over the last 50\textbf{??} years while Haddock has increased in abundance in recent years {[}{[}can cite the 2022 management track assessments, see \url{https://apps-nefsc.fisheries.noaa.gov/saw/sari.php} for when the document becomes available{]}{]} \textbf{just cite 2021 assessments?}. For this reason we compare survey index estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. We consider increasing, decreasing, and constant population trends for Yellowtail for completeness. The specific population trends used in our analyses can be see in Figure \ref{fig:pop-scenarios}. Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

\caption{True population trends used in index analyses. Spring biomass plots are shown with fall values being very similar.}\label{fig:pop-scenarios}
\end{figure}

\emph{Simulating Bottom Trawl Survey and Population Index}

After distributing each species in its stock area with simulated population trajectories, we mimic the bottom trawl survey by conducting stratified random sampling in each inhabited stratum twice each year. We sample each stratum in the same weeks in which the spring and fall surveys have historically sampled Georges Bank (weeks 13 \& 14 in the spring and 37 \& 38 in the fall). The number of simulated tows made reflects true target values for each stratum and sample cells where ``tows'' are made were randomly selected.

An underlying assumption in all index methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) consistently sampling the same proportion of the population. These assumptions can be questioned given enough noise in the sampling process and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare index estimates after adding noise to our samples versus those using the true sampling values. Annual survey observations were simulated as log-normal deviations from the underlying ``true'' survey catches with a CV of 0.3. The implication of a given population shifting its distribution into new habitat outside of the normal survey area is that stratified random sampling will fail to sample the entire geographic extent of the population. We simulate the effect of populations moving into new habitat by comparing index estimates using samples from all strata inhabited by each species on Georges Bank to those that only include a subset of the full spatial domain for each species. The strata to exclude for each species were chosen by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Figure \ref{fig:strata-plot} shows all strata inhabited by each species as well as those that are removed from certain calculations using the spatial shifting trends shown in Figure \ref{fig:PopPct}. The yellow regions in Figure \ref{fig:strata-plot} depict the strata that were not sampled for each species. We then use the biomass collected from our samples in contemporary abundance index methods to estimate yearly seasonal population trends.

The design-based method we used is the stratified mean, which divides the inhabited domain into \(N\) disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude. The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is proportional to the given area. The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated as the weighted average

\begin{equation*}
 SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i},
\end{equation*}

@@

We compare stratified mean estimates to those derived from the model-based VAST approach. VAST models both encounter probability \(p_1(i)\) and positive catch rate \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model that can be represented by

\[
  \makebox[\linewidth]{$f(p_*(i))= \beta_*(t_i) + L_{\omega_*}\omega_*(s_i) + L_* \epsilon_*(c_i,t_i) + \upsilon(s_i,t_i) + \xi(i),$}
\]

for \(*=1,2\), where more specific functional forms of each component are further described in Thorson (2019). VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved. As a result, VAST models take on the order of hours to complete.

@@

Knowing the true population values in our simulations allows us to compute the absolute error in a given season \(E_s\) from each estimation method and scenario as given by

\begin{equation}
E_s = \frac{\sum_{y=1}^{20}|M_{i,s}-I_{i,s}|}{\sum_{y=1}^{20}M_{i,s}},
\end{equation}

where \(M_{i,s}\) is the model biomass in year \(i\) and season \(s\) and \(I_{i,s}\) is the biomass estimate provided by either the stratified mean or VAST.

Through this process, and in consultation with the VAST package creator, we compared the performance of two model setting configurations in our VAST models as shown in Table 3. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online Github tutorial (Thorson, James T n.d.b). After running models for each of our scenarios using these settings, we received feedback to also consider configurations for Settings A (Thorson, James T n.d.c). Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings can be seen in the Rho Configuration, where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. The difference amounts to intercepts for the biomass predictor being a fixed effect (RhoBeta2=0) versus intercepts being constant among years as a fixed effect (RhoBeta2=3). The specifications for Settings A are more typical for stock assessment application (Thorson, James T n.d.c), and we therefore focus our discussion on these VAST results. Results for Settings B are shown to demonstrate how the settings used in model-based approaches impact model performance.

Each abundance index scenario we consider is a combination of specific population trends for each species, differing temperature scenarios, altering seasons, and sampling possibilities (noise, strata, \& covariates), resulting in a large number of scenario combinations to consider. The columns in Table \ref{tab:scenarios} show the choices that define each scenario.

@@

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the absolute error between biomass estimates and model output for each of our 20-year abundance estimates. While the model-based results provided a slightly lower overall mean absolute error (0.34) compared to the stratified mean (0.38), the variance of all VAST absolute errors (0.10) was much larger than the stratified mean (0.02). When considering individual scenarios, while about \(96\%\) had a VAST estimate with a lower absolute error than the corresponding stratified mean estimate, \(63\%\) of individual scenarios contained a VAST estimate with a worse error than the corresponding stratified mean estimate. VAST models that included covariate information provided the lowest overall absolute errors and standard deviations. Models with covariates had an average absolute error of 0.22, compared to models that did not include covariate information having an average error of 0.46, and stratified mean estimates produces an average error of 0.39. These findings demonstrate the ability of spatio-temporal models to account for spatial variation in the species being sampled under the right conditions, but also highlight volatility/sensitivity of model estimates to the settings being used and/or the covariate response detected in the data.

When we reduce the number of strata that are included in index calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain (e.g.~in Table \ref{tab:HadERROR} rows 8 and 9, VAST models without covariates show lower absolute errors than when covariates are included). We analyze these results further in the sections that follow.

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined through yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can result in yearly abundance values as low as 0 at times and exhibit large yearly changes (Figure \ref{fig:RatioTrend}).

Analysis of individual yearly estimate:model ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed similar patterns.

When the entire domain is sampled, the stratified mean produced an average yearly estimate:model ratio of 0.87 with a standard deviation of 0.12. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in turn decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower absolute error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.
